# Faster, Even Faster - with Package '*data.table*'

This repo is to illustrate how we can fasten our R code with *data.table* package.

We use *microbenchmark* to do the benchmarking for better accuracy.

```{r}
rm(list=ls())
library(data.table)
library(microbenchmark)

# Basic setting ---------------------------
set.seed(100)
N = 5e7L
benchmark_times <- 20
```

### How to subset a table faster 
```{r}
DT <- data.table(x = sample(letters, N, TRUE), 
                y = sample(1000L, N, TRUE), 
                val=runif(N), 
                key = c("x", "y")) # set the key
print(object.size(DT), units="Mb")
```

```{r}
microbenchmark(ans1 <- DT[x == "g" & y == 877L], 
               ans2 <- DT[.("g", 877L)],
               times = benchmark_times)

identical(ans1$val, ans2$val)
```





### How to update a table faster 

```{r}
DF <- data.frame(x = sample(letters, N, TRUE), 
                 y = sample(1000L, N, TRUE), 
                 val=runif(N))

DT <- as.data.table(DF)
```

##### Without Key (subsetting involved)
```{r}
microbenchmark(DF$y[DF$x == "x"] <- 0, 
               DT[x=="x", y := 0], 
               times = benchmark_times)
```

##### without Key (no subsetting involved)
```{r}
microbenchmark(DF$y <- 0, 
               DT[, y := 0], 
               times = benchmark_times)
```

##### With key (subsetting involved)
```{r}
setkey(DT, "x") # set the key
microbenchmark(DF$y[DF$x == "x"] <- 0, 
               DT[x=="x", y := 0], 
               DT[.("x"), y := 0],
               DT[.("x"), `:=`(y = 0)],
               times = benchmark_times)
```

##### With key (no subsetting involved)
```{r}
setkey(DT, "x") # set the key
microbenchmark(DF$y <- 0, 
               DT[, y := 0],
               DT[, `:=`(y = 0)],
               times = benchmark_times)
```

### How to Sort Your Table Faster
```{r}
# Generate the sample data.
DF <- data.frame(x = sample(letters, N, TRUE), 
                 y = sample(1000L, N, TRUE), 
                 val=runif(N))
DT <- as.data.table(DF)
```

```{r}
microbenchmark(DF <- DF[order(DF$x, DF$y),],
               DT <- DT[order(x, y)],
               times = benchmark_times)

```
